---
title: "Lab 08 - Text Mining/NLP"
output: github_document
always_allow_html: yes
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = F, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/
```{r eval=FALSE}
install.packages("tidytext")
install.packages("wordcloud2")
install.packages("tm")
install.packages("topicmodels")
```


```{r eval=TRUE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)
library("reshape2")

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r eval=FALSE}
mt_samples |>
  count(medical_specialty, sort = TRUE) |>
  ggplot(aes(fct_reorder(medical_specialty, n), n)) + 
  geom_col(fill ="dodgerblue") + 
  coord_flip() + 
  theme_bw()
```
There are 30 different medical specialties. The specialties are all pretty distinct, as all of these specialties do different tasks (even if the organs of the specialties are related to each other). These categories are not evenly distrubuted, as the most is seen in surgery. 
---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r eval=TRUE}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarize(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
tokens

tokens |>
  ggplot(aes(fct_reorder(word, word_frequency), word_frequency)) +
  geom_bar(stat="identity", fill="dodgerblue") +
  coord_flip() 

tokens |>
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.4, color ="random-light", backgroundColor = "black")
```
The results here don't really make sense, and it's because there are many stop-words (like and, the, this, etc.) that weren't filtered out. Hence, from this output, we don't really get any insightful insights. 
---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

```{r eval=FALSE}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

tokens2 <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription, token = "words") |>
  anti_join(stop_words, by="word") |>
  filter(!str_detect(word, "^[0-9]+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |>
  count(word, sort=TRUE) |>
  top_n(20, n)

tokens2 |>
  ggplot(aes(n, fct_reorder(word, n))) +
  geom_col(fill="dodgerblue") +
  theme_bw()

tokens2 |>
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.4, color ="random-light", backgroundColor = "black")

```
After filtering out the stop words, we get a much better understanding of what the text is about. We can see that it is medically related, specifically with keywords like patient, procedure, pain, anesthesia, blood etc. This gives us a much better insight of the text compared to having stop words being evaluated as well. 
---



## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r eval=FALSE}
stop_words2 <-c("mm", "mg", "noted", stop_words$word)
  
  
sw_start <- paste0("^", paste(stop_words2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stop_words2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!grepl(sw_start, ngram, ignore.case=TRUE))|>
  filter(!grepl(sw_end, ngram, ignore.case=TRUE))|>
  filter(!grepl("[[:digit:]]+", ngram))|>
  group_by(ngram) |>
  summarize(word_frequency=n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)

tokens_bigram |>
  ggplot(aes(ngram, word_frequency)) + 
  geom_col(fill="dodgerblue") + 
  coord_flip() 

```
Doing bigrams, we see a lot more common phrases, like "stable condition", "vital signs", which helps us further confirm that this text is medically related. We also see a lot of the words related to diagnoses, which lets us infer that this text has to do with confirming or finding diagnoses in patients. 
---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r eval=FALSE}
library(stringr)
# e.g. patient, blood, preoperative...
tokens_bigram |>
  filter(str_detect(ngram, regex("\\sblood$|^blood\\s"))) |> # finding paires with the word "blood" then we remove the word "blood"
  mutate(word = str_remove(ngram, "blood"),
  word = str_remove_all(word, " ")) |>
  # sum "xxx blod" and "blood xxx"
  group_by(word) |>
  head(20) |> 
  ggplot(aes(reorder(word, word_frequency), 
             word_frequency))+
  coord_flip() + 
  geom_col(fill = "dodgerblue")
```
Due to the long execution time from question 4 (ngrams), choosing the word `blood` didn't generate the top 20, but top 3 instead. 
---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r eval=FALSE}
mt_samples |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by ="word") |>
  filter(!str_detect(word, "^[0-9]+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |>
  group_by(medical_specialty) |>
  count(word, sort=TRUE) |>
  top_n(20, n)

```
In Surgery, I see that the words left, procedure, anesthesia, and incision are used a lot. In General Medicine, the word history is seen a lot. For Orthopedic, the word pain and left are seen the most.  However, in general, the word patient is the most used/seen word in each profession.  

## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r eval=FALSE}

transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by ="word") |>
  filter(!str_detect(word, "^[0-9]+$")) |>
  filter(!word %in% c("mm", "mg", "noted")) |>
  DocumentTermMatrix()

transcripts_dtm <- as.matrix(transcripts_dtm)   

transcripts_lda <- LDA(transcripts_dtm, k=5, control=list(seed=1234))

transcripts_lda
transcripts_top_terms <-
  tidy(transcripts_lda, matrix="beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |> 
  group_by(topic) |>
    slice_max(beta, n = 10) |>
  ungroup() |>
    arrange(topic, -beta)

transcripts_top_terms |>
  mutate(term=reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill=factor(topic))) + 
  geom_col(snow.legend=FALSE) + 
  facet_wrap(~topic, scales="free") +
  scale_y_reordered() +
  theme_bw()
```
A theme in the data is that the word pateint is in the top 3 words for all topics, which isn't surprising, since patient was the word that appeared the most. 



